A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-12-11 tex/codatadp.tex
More clarification on model.

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve terms containing $\mu$-bindings.
Recall that these bindings can be used for representing (possibly cyclic) values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $c$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if and only if it contains no free variables.
A $\mu$-term is \emph{cyclic} if and only if it contains a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some renaming of $\mu$-bound variables,
in which case we will commonly say $t$ and $s$ are \emph{bisimilar}.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.
First, we require additional terminology concerning $\mu$-terms.
If $u$ is a subterm of $\mu$-term $v$, then the \emph{interpretation of $u$ in $v$} is the $\mu$-term $v\interp{u}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subterm of $v$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{x}^M & = & x & \text{otherwise} \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M \\
v\interp{\mu x. C( \vec t )}^M & = & x & \text{otherwise}
\end{array}\]
We will write $v\interp{u}$ as shorthand for $v\interp{u}^\emptyset$.
Given a $\mu$-term $v$ of the form $\mu x. C( \vec t )$, 
we say $u$ is a \emph{bisimilar subterm} of $v$ if and only if $v\interp{u} \vsim v$.
We say $v$ is \emph{normal} if and only if all of its bisimilar subterms are $x$,
and all of its strict subterms that do not contain $x$ are normal.
For instance, the $\mu$-term $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is syntactically equivalent to $v$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.

%\begin{lemma}
%\end{lemma}

%For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
%(which we will call the \emph{normal form for $v$})
%that is observationally equivalent to $v$.
%We construct $\nf{v}$ by replacing all of the bisimilar subvalues of $v$ with $x$,
%and by replacing all of the non-normal closed subvalues of $v$ with their corresponding normal forms.
%For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.

@@

We construct a model $\M$ of $\thD$ that interprets each type $\tau$
as a maximal set of closed normal $\mu$-terms of that type that are 
(1) unique up to renaming of $\mu$-bound variables,
(2) acyclic when $\tau \not\in \Codata$.
We assume $\M$ interprets constructors and correctly-applied selectors in the obvious way
\footnote{
To interpret a correctly-applied selector, if $v = \mu x. C_i( \vec t )$, then $\M( s^j_i( v ) )$ is the value in $\M( \tau )$ that is a renaming of $v\interp{t_j}$,
where $t_j$ has type $\tau$.
},
interprets wrongly-applied selectors based on equality information in $\Fc$,
interprets the variables of $\Fc$ based on the valuation $\Val$ which was constructed in step two of our calculus.

@@

Let $\Val^\ast$ be the result of applying the above rule to completion on the $\Val$ we constructed during step 2 of our calculus.
We first argue that $FV( \Val^\ast ) = \emptyset$.
Since \rn{Split} does not apply to $\Fc$,
all equivalence classes of $\Fc$ finite datatype type contain constructor terms, 
and thus $FV( \Val )$ contains no variables of finite datatype type.
For each equivalence classes of $\Fc$ with finite co-datatype $\tau$, 
we know $\tau$ must have cardinality one, due to \rem{reference}.
Since \rn{Singleton} does not apply to $\Fc$, 
there is only one equivalence class of type $\tau$, call it $\ec{x}$.
%It cannot be the case that $v_{\ec{x}} \in FV(\Val^\ast \ec{x})$, then we may apply \rn{Complete} for the (single) value $c$ of type $\tau$, 
%for which the second premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.
If it were the case that $v_{\ec{x}} \in FV(\Val^\ast )$,
then \rn{Complete} would apply for the (single) value $c$ of type $\tau$, 
for which the second premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.
Since all $\tau \in \Nondata$ are infinite, all variables in $FV( \Val^\ast )$ have infinite type.
For any infinite type $\tau$,
there exists a closed normal $\mu$-term $c_1$ that is not bisimilar to any $c_2 \in \mathcal{T}(\Val^\ast)$, since $\mathcal{T}(\Val^\ast)$ is a finite set. \rem{elaborate?}
Thus, $FV( \Val^\ast )$ contains no variables of infinite type,
and consequently $\Val^\ast$ is a mapping from the equivalence classes of $\Fc$ to closed $\mu$-terms.
We now argue that:
\begin{eqnarray}
\label{eqn:m-normal} 
\Val^\ast \text{ maps to normal $\mu$-terms only},  \\
\label{eqn:m-bij} 
\Val^\ast \ec{x} \sim \Val^\ast \ec{y} \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}
We show this by induction on the number of applications of the rule \rn{Complete}.
For the base case, note that since \rn{Bisimilar} does not apply to $\Fc$,
we have that $\Val$ satisfies both~\ref{eqn:m-normal} and~\ref{eqn:m-bij}.
It satisfies the former since if a $\mu$-term $\mu v_{\ec{x}}. t_1$ had a bisimilar subterm $\mu v_{\ec{y}}. t_2$,
then the rule \rn{Bisimilar} would have applied to $x$ and $y$. \rem{elaborate?}
For the inductive case, say that we apply \rn{Complete} to $\Val$ for the substitution $\sigma := \{ v_{\ec{x}} \mapsto c_1 \}$.
\rem{TODO}

@@

If $\tau \in \Data$,
we first show that $\M( t ) = \Val^\ast \ec{x}$ for all non-selector terms $t$ that occur in datatype equivalence classes $\ec{x}$.
We show this by induction on the size of $\Val^\ast \ec{x}$.
In the case where $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t } = \Val^\ast \ec{ x }$.
In the case where $t$ is a constructor term $C( \vec u )$, 
then $\Val^\ast \ec{x}$ must have been assigned by \rn{Ind} for some constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val^\ast \ec{ x }$.

@@

If $\tau \in \Codata$, \rem{TODO}

>>>>

%For presentation of the rules in this step, we rely on a representation of terms in the $\mu$-notation.
The rules of this step contain premises that involve terms containing $\mu$-bindings.
Recall that these bindings can be used for representing (possibly cyclic) values,
for instance, $\mu x. \const{C}( \const{0}, x )$ represents the cyclic value $\const{C}( \const{0}, \const{C}( \const{0}, \ldots ))$.
Formally, we recursively define a $\mu$-term as a term-like structure either of the form
$\mu x. \const{C}( \vec t )$ for some constructor $\const{C} \in \Ctr$ and $\mu$-terms $\vec t$, 
or a variable $x$.
For consistency, we will also consider $\mu x. \const{c}$ to be a $\mu$-term
in the case where $c$ is a constant of type $\tau \in \Nondata$.
We consider only $\mu$-terms that are \emph{well-typed}, which we define recursively:
if the type of $\const{C}$ is $\vec \tau \rightarrow \upsilon$
and $\vec v$ are well-typed $\mu$-terms of type $\vec \tau$ under the assumption that $x$ has type $\upsilon$,
then $\mu x. \const{C}( \vec v )$ is a well-typed term of type $\upsilon$.
We say a $\mu$-term is \emph{closed} if and only if it contains no free variables.
A $\mu$-term is \emph{cyclic} if and only if it contains a bound variable.
We write $t \vsim s$ if $\mu$-terms $t$ and $s$ are syntactically equivalent for some renaming of $\mu$-bound variables,
in which case we will commonly say $t$ and $s$ are \emph{bisimilar}.
%, or informally are \emph{observationally equivalent}.
%For convienience, we will use $\mu$-terms to refer to (classes of) values for both co-datatype and datatype terms,
%where in the latter case, a $\mu$-term with a bound variable denotes an infeasible term.
%where the latter case adds the restriction on $\mu$-terms $t$ that no subterms of $t$ contain are bound variables.

@@

It remains to show the inverse of the previous lemma, that when a derivation tree with root node $\Ec$ contains a saturated node $\Fc$,
then $\Ec$ is satisfiable $\thD$.
First, we require additional terminology concerning $\mu$-terms.
If $u$ is a subterm of $\mu$-term $v$, then the \emph{interpretation of $u$ in $v$} is the $\mu$-term $v\interp{u}^\emptyset$ as returned by the following recursive procedure,
where $x \mapsto \mu x. C( \vec t ) \in v$ indicates that $\mu x. C( \vec t )$ is the subterm of $v$ that binds this occurrence of variable $x$.
\[\begin{array}{l@{\hspace{.5em}}l@{\hspace{.5em}}l@{\hspace{.5em}}l}
v\interp{x}^M & = & v\interp{\mu x. C( \vec t )}^M & x \not\in M, x \mapsto \mu x. C( \vec t ) \in v  \\
v\interp{x}^M & = & x & \text{otherwise} \\
v\interp{\mu x. C( \vec t )}^M & = & \mu x. C( v\interp{ \vec t }^{M \cup \{ x \}} ) & x \not\in M \\
v\interp{\mu x. C( \vec t )}^M & = & x & \text{otherwise}
\end{array}\]
We will write $v\interp{u}$ as shorthand for $v\interp{u}^\emptyset$.
Given a $\mu$-term $v$ of the form $\mu x. C( \vec t )$, 
we say $u$ is a \emph{bisimilar subterm} of $v$ if and only if $v\interp{u} \vsim v$.
We say $v$ is \emph{normal} if and only if all of its bisimilar subterms are $x$,
and all of its strict subterms are also normal.
For instance, the $\mu$-term $v$ of the form $\mu x. \const{S}( \mu y. \const{S}( x ) )$ is not normal, 
since 
$v\interp{\mu y. \const{S}( x )}^\emptyset =$ 
$\mu y. \const{S}( v\interp{x}^{\{y\}} ) =$ 
$\mu y. \const{S}( v\interp{\mu x. \const{S}( \mu y. \const{S}( x ) ) }^{\{y\}} ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( v\interp{\mu y. \const{S}( x )}^{\{x,y\}} ) ) =$ 
$\mu y. \const{S}( \mu x. \const{S}( y ) )$,
which is syntactically equivalent to $v$ for the renaming $\{ x \mapsto y, y \mapsto x \}$.
For any $\mu$-value $v$ of the form $\mu x. C( \vec w )$, we may recursively construct a normal value $\nf{v}$
(which we will call the \emph{normal form} of $v$)
that is a renaming of $v$ by replacing all of the bisimilar subvalues of $v$ with $x$,
and by replacing all of the non-normal subterms of $v$ with their corresponding normal forms.
For instance, $\nf{(\mu x. \const{S}( \mu y. \const{S}( x ) ))} = \mu x. \const{S}( x )$.

@@

We construct a model $\M$ of $\thD$ that interprets each type $\tau$
as a maximal set of closed normal $\mu$-terms of that type that are 
unique up to renaming of $\mu$-bound variables,
and acyclic when $\tau \not\in \Codata$.
The interpretation of constructor terms
$\M( C( \vec t ) )$ for $C \in \Ctr$ is the value in $\M( \tau )$ that is a renaming of $\nf{(\mu x. C( \M( \vec t ) ))}$ where $x \not\in FV( \vec t )$.
The interpretation of (correctly-applied) selector terms
$\M( s^j_i( t ) )$ for $s^j_i \in \Sel$ is the value in $\M( \tau )$ that is a renaming of $\M( t )\interp{u_j}$
where $\M( t )$ is of the form $\mu x. C_i( \vec u )$ for some $\vec u$.
For instance, if $\M( t ) = \mu x. \const{C}_1( x )$ (call this term $v$), 
then $\M( s_1^1( t ) )$ is a renaming of
$v\interp{x} =$ 
$v\interp{\mu x. \const{C}_1( x )}^\emptyset =$
$\mu x. \const{C}_1( v\interp{x}^{\{x\}} ) =$
$\mu x. \const{C}_1( x )$.
%To interpret a correctly-applied selector in $\M$, if $v = \mu x. C_i( \vec t )$, then $\M( s^j_i( v ) )$ is the value in $\M( \tau )$ that is a renaming of $v\interp{t_j}$,
%where $t_j$ has type $\tau$.
%We assume $\M$ interprets constructors and correctly-applied selectors in the obvious way
%\footnote{
%The interpretation of $\M( C( \vec t ) )$ is the value in $\M( \tau )$ that is a renaming of $\nf{\mu x. C( \M( \vec t ) ) }$.
%To interpret a correctly-applied selector, if $v = \mu x. C_i( \vec t )$, then $\M( s^j_i( v ) )$ is the value in $\M( \tau )$ that is a renaming of $v\interp{t_j}$,
%where $t_j$ has type $\tau$.
%},
Additionally, $\M$ will interpret wrongly-applied selectors based on equality information in $\Fc$,
and will interpret the variables of $\Fc$ based on the valuation $\Val$ which was constructed in step two of our calculus.

@@

Let $\Val^\ast$ be the result of applying the above rule to completion on the $\Val$ we constructed during step 2 of our calculus.
We first argue that $FV( \Val^\ast ) = \emptyset$.
Since \rn{Split} does not apply to $\Fc$,
all equivalence classes of $\Fc$ finite datatype type contain constructor terms, 
and thus $FV( \Val )$ contains no variables of finite datatype type.
For each equivalence classes of $\Fc$ with finite co-datatype $\tau$, 
we know $\tau$ must have cardinality one, due to \rem{reference}.
Since \rn{Singleton} does not apply to $\Fc$, 
there is only one equivalence class of type $\tau$, call it $\ec{x}$.
%It cannot be the case that $v_{\ec{x}} \in FV(\Val^\ast \ec{x})$, then we may apply \rn{Complete} for the (single) value $c$ of type $\tau$, 
%for which the second premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.
If it were the case that $v_{\ec{x}} \in FV(\Val^\ast )$,
then \rn{Complete} would apply for the (single) value $c$ of type $\tau$, 
for which the second premise holds since $\ec{x}$ is the only equivalence class of type $\tau$.
Since all $\tau \in \Nondata$ are infinite, all variables in $FV( \Val^\ast )$ have infinite type.
For any infinite type $\tau$,
there exists a closed normal $\mu$-term $c_1$ that is not bisimilar to any $c_2 \in \mathcal{T}(\Val^\ast)$, 
since $\mathcal{T}(\Val^\ast)$ is a finite set (noting that a closed $\mu$-term is only bisimilar to other closed $\mu$-terms).
Thus, $FV( \Val^\ast )$ contains no variables of infinite type,
and consequently $\Val^\ast$ is a mapping from the equivalence classes of $\Fc$ to closed $\mu$-terms.
We now argue that:
\begin{eqnarray}
\label{eqn:m-normal} 
\Val^\ast \text{ maps to normal $\mu$-terms only},  \\
\label{eqn:m-bij} 
\Val^\ast \ec{x} \sim \Val^\ast \ec{y} \text{ if and only if } \ec{x} = \ec{y}
\end{eqnarray}
We show this by induction on the number of applications of the rule \rn{Complete}.
For the base case, note that since \rn{Bisimilar} does not apply to $\Fc$,
we have that $\Val$ satisfies both~\ref{eqn:m-normal} and~\ref{eqn:m-bij}.
It satisfies the former since if a $\mu$-term $\mu v_{\ec{x}}. t_1$ had a bisimilar subterm $\mu v_{\ec{y}}. t_2$,
then the rule \rn{Bisimilar} would have applied to $x$ and $y$. \rem{elaborate?}
For the inductive case, we assume $\Val$ satisfies ~\ref{eqn:m-normal} and~\ref{eqn:m-bij},
and show that $\Val \sigma$ does as well where $\sigma := \{ v_{\ec{x}} \mapsto c_1 \}$ is a substitution applied by \rn{Complete}.
By contradiction, assume that $\Val \sigma \ec{y}$ is a non-normal $\mu$-term for some $\ec{y}$.
%Since $c_1$ is normal, we have $\ec{x} \neq \ec{y}$, 
%and thus $\Val \ec{y}$ must have been of the form $\mu v_{\ec{y}}. t$ where $\ec{x} \in FV( t )$.
Let $u$ be the (strict) subterm of $\Val \ec{y}$ such that $(\Val \ec{y} \sigma ) \interp{u \sigma } \sim (\Val \ec{y} \sigma)$. 
Since $\Val$ maps only to normal $\mu$-terms, we have $x \in FV( u )$.
\rem{TODO}

@@

If $\tau \in \Data \cup \Codata$,
we first show that $\M( t ) = \Val^\ast \ec{x}$ for all non-selector terms $t$ that occur in equivalence classes $\ec{x}$.
We show this by induction on the size of $\Val^\ast \ec{x}$.
In the case where $t$ is a variable, then $\M( t ) = \Val^\ast \ec{ t } = \Val^\ast \ec{ x }$.
In the case where $t$ is a constructor term $C( \vec u )$, 
then $\Val^\ast \ec{x}$ must have been assigned by \rn{Ind} for some constructor term $C'( \vec u' ) \in \ec{x}$.
Since \rn{Unify$_2$} does not apply to $\Fc$, we have that $C = C'$.
Since \rn{Unify$_1$} does not apply to $\Fc$, we have that $\ec{\vec u} = \ec{\vec u'}$.
By the induction hypothesis, we have $\M( \vec u ) = \M( \vec u' )$, and thus $\M( t ) = C( \M( \vec u ) ) = C'( \M( \vec u ' ) ) = \Val^\ast \ec{ x }$.

@@
